---
title: "Validation breast cancer metagene score"
author: "M. Hoogstraat"
date: "3/24/2022"
output: html_document
---

# R code to validate the metagene score from M. Hoogstraat et al., in an independent dataset

The new dataset should contain normalized gene expression data in log scale. 
These values can then be transformed using the (modified) Z-score, for example:
```{r z_transform_example, eval=F}
exprs_z <- 0.6745 * (exprs - apply(exprs, 1, median)) / abs(apply(exprs, 1, mad))
```
...where 'exprs' is a dataframe containing ONLY the normalized, log scaled expression data, with samples in columns and genes in rows.
Column headers should contain sample identifiers.
Rownames should contain gene identifiers (HUGO symbol, ensembl gene id, entrez id).

As an example, the R source code to validate the signature in the TCGA breast cancer dataset is provided below.

# Validation of the metagene score in the [TCGA BRCA 2018 dataset](https://www.sciencedirect.com/science/article/pii/S0092867418303027)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Loading required libraries
```{r loading_libs, message=F, warning=F}
library(survival)
```

#### Loading required files
```{r loading_data}
rna_zscores <- read.delim("data_RNA_Seq_v2_mRNA_median_Zscores.txt") #Source data not provided
clindat <- read.delim("data_clinical_patient.txt",comment.char="#") #Source data not provided
clindat$colnames <- paste(gsub("-",".",clindat$PATIENT_ID),".01",sep="")
metagene_genes<-read.delim("metagene_genelist.txt")
rownames(metagene_genes) <- metagene_genes$gene_name
```

#### Testing the amount of genes from original signature present in new data
```{r test_overlap}
knitr::kable(table(metagene_genes$gene_name[which(metagene_genes$process %in% c('proliferation','immune response','ECM associated'))] %in% rna_zscores$Hugo_Symbol), caption="Genes from signature present in new data")
```

## BRCA_Basal

#### Obtain gene expression levels from signature genes
```{r get_genes_Basal}
rna_zscores_basal <- as.data.frame(rbind(
	rna_zscores[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "proliferation")]),
		as.vector(clindat$colnames[which(clindat$SUBTYPE == "BRCA_Basal")])],
	rna_zscores[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "immune response")]),
		as.vector(clindat$colnames[which(clindat$SUBTYPE == "BRCA_Basal")])]
	))

rna_zscores_basal <- as.data.frame(rbind(
	rna_zscores_basal,
	rna_zscores[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "ECM associated")]),
		as.vector(clindat$colnames[which(clindat$SUBTYPE == "BRCA_Basal")])]
	))

rownames(rna_zscores_basal) <- c(
	as.vector(rna_zscores$Hugo_Symbol[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "proliferation")])]),
	as.vector(rna_zscores$Hugo_Symbol[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "immune response")])]),
	as.vector(rna_zscores$Hugo_Symbol[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "ECM associated")])])
)
```

#### Calculate metagenes and cutoffs
```{r get_metagenes_Basal}
clindat_basal <- clindat[which(clindat$SUBTYPE == "BRCA_Basal"),]

clindat_basal$Proliferation <- apply(rna_zscores_basal[as.vector(metagene_genes$gene_name[which(metagene_genes$process == "proliferation" & metagene_genes$gene_name %in% rownames(rna_zscores_basal))]),],2,median)

clindat_basal$Immune <- apply(rna_zscores_basal[as.vector(metagene_genes$gene_name[which(metagene_genes$process == "immune response" & metagene_genes$gene_name %in% rownames(rna_zscores_basal))]),],2,median)

clindat_basal$ECM <- apply(rna_zscores_basal[as.vector(metagene_genes$gene_name[which(metagene_genes$process == "ECM associated" & metagene_genes$gene_name %in% rownames(rna_zscores_basal))]),],2,median)

clindat_basal$cutoffs <- "other"
clindat_basal$cutoffs[which(clindat_basal$Proliferation >= median(clindat_basal$Proliferation) & clindat_basal$Immune >= median(clindat_basal$Immune) & clindat_basal$ECM < median(clindat_basal$ECM))] <- "HHL"
clindat_basal$cutoffs[which(clindat_basal$Proliferation < median(clindat_basal$Proliferation) & clindat_basal$Immune < median(clindat_basal$Immune) & clindat_basal$ECM >= median(clindat_basal$ECM))] <- "LLH"
```

#### Process clinical data
```{r parse_clindat_Basal}
clindat_basal$PFS_stat_bin <- 0
clindat_basal$PFS_stat_bin[which(clindat_basal$PFS_STATUS == "PROGRESSION")] <- 1

clindat_basal$n_stage <- 1
clindat_basal$n_stage[grep("N0",clindat_basal$PATH_N_STAGE)] <- 0
clindat_basal$t_stage_bin <- NA
clindat_basal$t_stage_bin[grep("T[12]",clindat_basal$PATH_T_STAGE)] <- "1_2"
clindat_basal$t_stage_bin[grep("T[34]",clindat_basal$PATH_T_STAGE)] <- "3_4"
```

#### Calculate association metagene score vs progression-free survival in various subgroups
```{r associate_clindat_metagene_Basal}
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal)
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$n_stage == 1 | clindat_basal$t_stage_bin == "3_4"),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$n_stage == 0 ),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$n_stage == 1 ),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$t_stage_bin == "1_2"),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$t_stage_bin == "3_4"),])
```

#### Visualize association metagene score vs progression-free survival in various subgroups
```{r visualize_assoc_Basal, fig.height=10}
par(mfrow=c(3,2))
plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal), mark.time=T,col=c("orange","blue","grey"),main="All_basal",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,20))
legend('bottomleft',lwd=2,col=c('blue','orange','grey'), legend=c('HHL','LLH','other'),cex=1.5)
text(15,0.2,labels="P = 0.2",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$n_stage == 1 | clindat_basal$t_stage_bin == "3_4"),]), mark.time=T,col=c("orange","blue","grey"),main="Basal, N+ or T-stage 3 or 4",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,20))
text(15,0.2,labels="P = 0.02",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$n_stage == 0 ),]), mark.time=T,col=c("orange","blue","grey"),main="Basal, N-",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,20))
text(15,0.2,labels="P = 0.7",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$n_stage == 1 ),]), mark.time=T,col=c("orange","blue","grey"),main="Basal, N+",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,20))
text(15,0.2,labels="P = 0.004",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$t_stage_bin == "1_2" ),]), mark.time=T,col=c("orange","blue","grey"),main="Basal, T-stage 1 or 2",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,20))
text(15,0.2,labels="P = 0.8",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_basal[which(clindat_basal$t_stage_bin == "3_4" ),]), mark.time=T,col=c("orange","blue","grey"),main="Basal, T-stage 3 or 4",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,20))
text(15,0.2,labels="P = 0.3",cex=2)
```

## BRCA_Luminal

#### Obtain gene expression levels from signature genes
```{r get_genes_Lum}
rna_zscores_lum <- as.data.frame(rbind(
	rna_zscores[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "proliferation")]),
		as.vector(clindat$colnames[which(clindat$SUBTYPE %in% c("BRCA_LumA", "BRCA_LumB"))])],
	rna_zscores[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "immune response")]),
		as.vector(clindat$colnames[which(clindat$SUBTYPE %in% c("BRCA_LumA", "BRCA_LumB"))])]
	))
rna_zscores_lum <- as.data.frame(rbind(
	rna_zscores_lum,
	rna_zscores[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "ECM associated")]),
		as.vector(clindat$colnames[which(clindat$SUBTYPE %in% c("BRCA_LumA", "BRCA_LumB"))])]
	))

rownames(rna_zscores_lum) <- c(
	as.vector(rna_zscores$Hugo_Symbol[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "proliferation")])]),
	as.vector(rna_zscores$Hugo_Symbol[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "immune response")])]),
	as.vector(rna_zscores$Hugo_Symbol[which(rna_zscores$Hugo_Symbol %in% metagene_genes$gene_name[which(metagene_genes$process == "ECM associated")])])
)
```

#### Calculate metagenes and cutoffs
```{r get_metagenes_Lum}
clindat_lum <- clindat[which(clindat$SUBTYPE %in% c("BRCA_LumA", "BRCA_LumB")),]
clindat_lum$PFS_stat_bin <- 0
clindat_lum$PFS_stat_bin[which(clindat_lum$PFS_STATUS == "PROGRESSION")] <- 1

clindat_lum$Proliferation <- apply(rna_zscores_lum[as.vector(metagene_genes$gene_name[which(metagene_genes$process == "proliferation" & metagene_genes$gene_name %in% rownames(rna_zscores_lum))]),],2,median)
clindat_lum$Immune <- apply(rna_zscores_lum[as.vector(metagene_genes$gene_name[which(metagene_genes$process == "immune response" & metagene_genes$gene_name %in% rownames(rna_zscores_lum))]),],2,median)
clindat_lum$ECM <- apply(rna_zscores_lum[as.vector(metagene_genes$gene_name[which(metagene_genes$process == "ECM associated" & metagene_genes$gene_name %in% rownames(rna_zscores_lum))]),],2,median)
```

#### Process clinical data
```{r parse_clindat_Lum}
clindat_lum$cutoffs <- "other"
clindat_lum$cutoffs[which(clindat_lum$Proliferation >= median(clindat_lum$Proliferation) & clindat_lum$Immune >= median(clindat_lum$Immune) & clindat_lum$ECM < median(clindat_lum$ECM))] <- "HHL"
clindat_lum$cutoffs[which(clindat_lum$Proliferation < median(clindat_lum$Proliferation) & clindat_lum$Immune < median(clindat_lum$Immune) & clindat_lum$ECM >= median(clindat_lum$ECM))] <- "LLH"

clindat_lum$n_stage <- 1
clindat_lum$n_stage[grep("N0",clindat_lum$PATH_N_STAGE)] <- 0
clindat_lum$t_stage_bin <- NA
clindat_lum$t_stage_bin[grep("T[12]",clindat_lum$PATH_T_STAGE)] <- "1_2"
clindat_lum$t_stage_bin[grep("T[34]",clindat_lum$PATH_T_STAGE)] <- "3_4"
```

#### Calculate association metagene score vs progression-free survival in various subgroups
```{r associate_clindat_metagene_Lum}
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum)
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$n_stage == 1 | clindat_lum$t_stage_bin == "3_4"),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$n_stage == 0 ),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$n_stage == 1 ),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$t_stage_bin == "1_2" ),])
survdiff(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$t_stage_bin == "3_4" ),])
```

#### Visualize association metagene score vs progression-free survival in various subgroups
```{r visualize_assoc_Lum, fig.height=10}
par(mfrow=c(3,2))
plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum), mark.time=T,col=c("orange","blue","grey"),main="All_lum",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,14))
legend('bottomleft',lwd=2,col=c('blue','orange','grey'), legend=c('HHL','LLH','other'),cex=1.5)
text(10,0.2,labels="P = 0.09",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$n_stage == 1 | clindat_lum$t_stage_bin == "3_4"),]), mark.time=T,col=c("orange","blue","grey"),main="Luminal, N+ or T-stage 3 or 4",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,14))
text(10,0.2,labels="P = 0.1",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$n_stage == 0 ),]), mark.time=T,col=c("orange","blue","grey"),main="Luminal, N-",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,14))
text(10,0.2,labels="P = 0.8",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$n_stage == 1 ),]), mark.time=T,col=c("orange","blue","grey"),main="Luminal, N+",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,14))
text(10,0.2,labels="P = 0.1",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$t_stage_bin == "1_2" ),]), mark.time=T,col=c("orange","blue","grey"),main="Luminal, T-stage 1 or 2",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,14))
text(10,0.2,labels="P = 0.5",cex=2)

plot(survfit(Surv(PFS_MONTHS/12, PFS_stat_bin) ~ cutoffs, data=clindat_lum[which(clindat_lum$t_stage_bin == "3_4" ),]), mark.time=T,col=c("orange","blue","grey"),main="Luminal, T-stage 3 or 4",xlab="Progression-free survival (years)",ylab="Proportion of patients at risk",lwd=2, cex.lab=1.5, cex.axis=1.5,cex.main=1.5,xlim=c(0,14))
text(10,0.2,labels="P = 0.1",cex=2)
```


